Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-10-12 Manuscript/manuscript.tex
New abstract and small changes in the text

\begin{abstract}
  % insert abstract here
  \todo{Abstract directly from Joe's conference abstracts. To be rewritten.}
Classical molecular dynamics simulations give detailed information about membrane structure and dynamics. 
However, there is still a room for improvements in current force fields – it is known from the literature,
that the binding of ions, especially cations, to phopholipid membranes is overestimated in all classical models [1]. 
We suggest that the membrane-ion interactions can be corrected by including implicit electronic polarizability into the
lipid models through the electronic continuum correction (ECC) [2], which was already applied to monovalent and divalent
ions yielding models that feature correct ion pairing [3]. 
Using the electrometer concept [3, 4] and x-ray scattering form factors, our simulations point out that our hypothesis is
correct and ECC is indeed a missing important contribution in current classical lipid models. 
Moreover, the solid physical principles behind ECC are found not to hamper other relevant properties of a phospholipid bilayer. 
The new lipid model, "ECC-lipids", shows accurate binding affinity to sodium and calcium cations and head group order parameter response to bound charge. 
We also provide for the first time a realistic stochiometry of bound calcium cations to a POPC membrane, and their binding sites. 
This work will continue as an open collaboration project NMRlipids VI (http://nmrlipids.blogspot.fi).
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. Especially the zwitterionic phosphocholine (PC) lipid bilayers are
often used as model systems to understand the role of ions in complex biological
systems~\cite{scherer87,seelig90,cevc90}.

@@

The concept can be used to compare the ion binding affinity in lipid 
bilayers between MD simulations and  NMR experiments, because the order
parameters can be accurately determined from both techniques~\cite{ollila16}.
The order parameters for all C-H bonds in lipid molecules, including
$\alpha$ and $\beta$ segments in head group, can be accurately measured
using $^2$H NMR or $^{13}$C NMR techniques. From MD simulations the
order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and the average is taken over all sampled configurations \cite{ollila16}.

@@

\subsection{Salt concentrations and binding affinity}
The early experimental electrometer concept data for the lipid head group
order parameter changes was reported as a function 
of the salt concentration in water before solvating
the lipids~\cite{akutsu81}. The later study used atomic absorption
spectroscopy and reported the order parameter changes as
a function of the salt concentration in the supernatant after the
solvation of lipids~\cite{altenbach84}.
In this work we focus on POPC for which the latter definition was used~\cite{altenbach84}.
The salt concentration in the aqueous bulk region was calculated from 
the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study, Ref.~\citenum{catte16},
the ion concentrations were calculated in water before solvating the lipids 
as in the earlier experiments~\cite{akutsu81}.
Despite of the measurable differences between these two concentrations for CaCl$_2$ systems,
the qualitative conclusions in this or in the previous work~\cite{catte16} are
not affected.

@@

The structure of lipid bilayers in simulations without ions were validated against NMR and 
\mbox{x-ray} scattering experiments by calculating order parameters
for C-H bonds and form factors. The former validates the structures sampled by individual
lipid molecules in simulations with atomic resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}.

@@

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium.%, which is one of the cationic surfactants used to experimentally quantify the electrometer concept~\cite{scherer89}.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The parameters are available at \cite{??} \todo{To be added}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. Chloride counter ions were used
in simulations, because parameters for bromide, the counterion in the experimental data \cite{scherer89},
were not available in the standard Gromacs files for Amber force field.
The Lipid14 model was used for POPC.
The first 20~ns of the total simulation time of 200~ns
was considered as an equlibration time and was omitted from the analysis.
A reasonable lipid neighbor exchange occured during the simulation.

The same systems were also simulated with the ECC-lipid model for POPC using the same setup.
\todo{Which water model was used in these simulations?}
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all charges with the same factor as for ECC-lipids, i.e., $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids.

@@

\begin{table}
  \caption{Area per lipid (APL) from different models of POPC with no ions\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & 62.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{jambeck12}\todoii{REF}{put original references, not Slipids param. paper.} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}

>>>>

\begin{abstract}
  % insert abstract here
  Despite of the significant biological relevance, the binding details of cations in cellular
  lipid membranes are not fully understood. For example, consistent results for
  the binding affinities and stoichimetries of Na$^+$ and Ca$^{2+}$ ions to phospholipid bilayers have not been
  achieved with different experimental and theoretical methods. As recently shown in the 
  NMRlipids project, a Open Collaboration running at \url{nmrlipids.blogspot.com},
  the ion binding details could be resolved with unprecendent detail by intepreting
  the experimental NMR data with classical molecular dynamics (MD) simulations.
  However, the accuracy of the existing lipid models for MD simulations was not sufficient for this.
  In this work we show that the binding details of Na$^+$ and Ca$^{2+}$ ions to
  1-Palmitoyl-2-oleoylphosphatidylcholine (POPC) bilayer
  can be accurately described with a MD simulation model having implicitly included electronic polarization.
  This is demonstrated by applying the electronic continuum correction (ECC) to a
  state of the art lipid model for MD simulations of POPC
  lipid bilayer. The introduced ECC-lipid model reproduces the experimentally measured structural parameters for the ion-free membrane,
  the response of lipid headgroup to the bound positive charge, and the binding affinities of
  Na$^+$ and Ca$^{2+}$ ions. The imperceptible binding of Na$^+$ ions to POPC bilayer
  and interactions of Ca$^{2+}$ mainly with phoshpate oxygens in the ECC-lipid model give
  support to the early intepretations of the experimental spectroscopic data.
  On the other hand, Ca$^{2+}$ ions form complexes with 1-3 lipid molecules with almost
  equal probabilities, suggesting more complicated binding stoichimetry than 
  the simple binding model used to interpret the NMR data.
  The results in this work pave the way for MD simulations of complex biochemical systems
  with realistic electrostatic interactions in the vicinity of cellular membranes.
\end{abstract}

@@

% Here I write pseudo-article statements that will make the main argument.
% Beautiful polished sentences will be formed only afte we agree on these basic things
\section{Introduction}
Cation interactions with cellular membranes play a key role in several biological processes,
such as in signal propagation in neurons and vesicle fusion.
Since the direct measurements of ion-membrane interactions from biological systems
are difficult, lipid bilayers are often used as models systems for cellular
membranes. Especially the zwitterionic phosphocholine (PC) lipid bilayers are
used to understand the role of ions in complex biological
systems~\cite{scherer87,seelig90,cevc90}.

@@

The concept can be used to compare the ion binding affinity in lipid 
bilayers between MD simulations and  NMR experiments, because the order
parameters can be accurately determined from both techniques~\cite{catte16,ollila16}.
The order parameters for all C-H bonds in lipid molecules, including
$\alpha$ and $\beta$ segments in head group, can be accurately measured
using $^2$H NMR or $^{13}$C NMR techniques \cite{ollila16}. From MD simulations the
order parameters can be calculated using the definition
\begin{equation}\label{OP}
S_{\rm CH}=\frac{3}{2}\langle \cos^2\theta -1 \rangle,
\end{equation}
where $\theta$ is the angle between the bond and membrane
normal and the average is taken over all sampled configurations \cite{ollila16}.

@@

\subsection{Salt concentrations and binding affinity}
The early experimental electrometer concept data for the lipid head group
order parameter changes was reported as a function 
of the salt concentration in water before solvating
the lipids~\cite{akutsu81}. The later study used atomic absorption
spectroscopy and reported the order parameter changes as
a function of the salt concentration in the supernatant after the
solvation of lipids~\cite{altenbach84}.
In this work we focus on POPC for which the latter definition was used~\cite{altenbach84}.
The salt concentration in the aqueous bulk region was calculated from 
the farthest point from both lipid leaflets in the water phase. 
Note that in the previous study, Ref.~\citenum{catte16},
the ion concentrations were calculated in water before solvating the lipids 
as in the earlier experiments~\cite{akutsu81}.
Despite of the measurable differences between these two
definitions of concentrations for CaCl$_2$ systems,
the qualitative conclusions in this or in the previous work~\cite{catte16} are
not affected by this.

@@

The structure of lipid bilayers in simulations without ions were validated against NMR and 
\mbox{x-ray} scattering experiments by calculating the order parameters
for C-H bonds and the scattering form factors. The former validates the structures sampled by individual
lipid molecules in simulations with atomic resolution, while the latter
validates the dimensions of the lipid bilayer (thickness and area per
molecule)~\cite{ollila16}.

@@

\subsubsection{Simulations of POPC bilayers with cationic surfactants}
An automated topology builder \cite{malde11} was first used to create the structure of
dihexadecyldimethylammonium.
The AmberTools program \cite{amber} was then used to generate the Amber-type force field
parameters. The parameters were converted to the Gromacs format by using
the acpype tool \cite{acpype}. The partial charges were then manually modified
to approximately correspond to their equivalent segments in Lipid14 \cite{dickson14}.
The parameters are available at \cite{??} \todo{To be added}.
The surfactants were randomly placed among the lipids to form bilayer structures with
mole fractions of 10\%, 20\%, 30\%, 42\%, or 50\% of surfactant in the POPC bilayer.
All systems contained 50 POPC molecules per leaflet, 6340 TIP3P water molecules and
6, 14, 21, 35, or 50 surfactants per leaflet. Chloride counter ions were used
in simulations, because parameters for bromide, the counterion in the experimental data \cite{scherer89},
were not available in the standard Gromacs files for Amber force field.
The Lipid14 model was used for POPC.
The first 20~ns of the total simulation time of 200~ns
was considered as an equlibration time and was omitted from the analysis.
A reasonable lipid neighbor exchange occured during the simulation.

The same systems were also simulated with the ECC-lipid model for POPC using the same setup.
\todo{Which water model was used in these simulations?}
In these simulations the ECC correction was also applied to the cationic surfactant 
by scaling all charges with the same factor as for ECC-lipids, i.e., $f_q=0.8$, 
and by using the atom types with reduced $\sigma$ parameters from ECC-lipids. 
\todo{Parameters are available at ??}

@@

\begin{table}
  \caption{Area per lipid (APL) values of POPC with no ions from the Lipid14 simulation
    ran in this work and from the literature, from ECC-lipid model and from experiments.\label{tab:apls} }
  \begin{tabular}{l|c c}
    model          & APL (\AA$^2$)   & Temperature [K] \\
    \hline
    Lipid14                   & 65.1$\pm$ 0.6  &  300 \\
    Lipid14 \cite{dickson14}  & 65.6$\pm$ 0.5  &  303 \\
    \hline
    ECC-lipids                & 62.2$\pm$ 0.6  &  300       \\
    \hline
    experiment \cite{jambeck12}\todoii{REF}{put original references, not Slipids param. paper.} & 64.3  &  303    \\
    \hline
  \end{tabular}
\end{table}
